---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
---

## Synopsis

Per capita income and expenditures provide crucial insight into the average standard of living in specified areas. Disposable per capita income measures the average income earned after taxes per person in a given area (city, state, country, etc.) in a specified year. It is calculated by dividing the area's total income after tax by its total population. Per capita expenditures, on the other hand, measures the average outlay for goods and services by person and provides insight into spending patterns across a given area.  Together, the assessment of per capita income versus expenditures can provide better understanding of regional economies, differences in standard of living, and approximate savings rates.

This project involves exploring [Bureau of Economic Analysis](http://www.bea.gov/index.htm) data regarding [per capita disposable personal income](http://www.bea.gov/iTable/iTableHtml.cfm?reqid=70&step=30&isuri=1&7022=21&7023=0&7024=non-industry&7033=-1&7025=0&7026=00000,01000,02000,04000,05000,06000,08000,09000,10000,11000,12000,13000,15000,16000,17000,18000,19000,20000,21000,22000,23000,24000,25000,26000,27000,28000,29000,30000,31000,32000,33000,34000,35000,36000,37000,38000,39000,40000,41000,42000,44000,45000,46000,47000,48000,49000,50000,51000,53000,54000,55000,56000&7027=-1&7001=421&7028=53&7031=0&7040=-1&7083=levels&7029=23&7090=70) (hereafter referred to as PCI) and [per capita personal expenditures](http://www.bea.gov/iTable/iTableHtml.cfm?reqid=70&step=10&isuri=1&7003=2&7035=-1&7004=x&7005=1&7006=00000,01000,02000,04000,05000,06000,08000,09000,10000,11000,12000,13000,15000,16000,17000,18000,19000,20000,21000,22000,23000,24000,25000,26000,27000,28000,29000,30000,31000,32000,33000,34000,35000,36000,37000,38000,39000,40000,41000,42000,44000,45000,46000,47000,48000,49000,50000,51000,53000,54000,55000,56000&7036=-1&7001=62&7002=6&7090=70&7007=-1&7093=levels) (hereafter referred to as PCE). The PCI data provides annual (non-inflation adjusted) per capita disposable income at the national and state-level from 1948-2015 and the PCE data provides annual (non-inflation adjusted) per capita personal consumption expenditures at the national and state-level from 1997-2014. Consequently, this research seeks to identify how the national and state-level savings rates defined as $Savings = PCI - PCE$ has changed over time and by geographic location.

The analysis finds that the savings rate has increased/decreased from XX% in 1997 to XX% in 2014. This was primarily driven by XXXX as disposable income has only grown by XX% while expenditures have grown by XX%.  Furthermore, we find that significant differences exist at the state level. *Add some more explanation here...*


## Packages Required

To reproduce the code and results throughout this project you will need to load the following packages.

```{r, message=FALSE, warning=FALSE}
library(rvest)        # scraping data
library(tidyr)        # creating tidy data
library(dplyr)        # transforming (joining, summarizing, etc.) data
library(tibble)       # coercing data to tibbles
library(magrittr)     # for piping capabilities
library(DT)           # for printing nice HTML output tables
library(ggplot2)      # visualizing data
library(ggrepel)      # Repel overlapping text labels in plots
```


## Data Preparation

Prior to assessing how PCI, PCE, and savings rates have behaved over time and by geographic location we must acquire and clean the data.

### Loading Data

The data for this project originated from the following sources:

- PCI data: [http://bit.ly/2dpEPY1](http://bit.ly/2dpEPY1) 
- PCE data: [http://bit.ly/2dhC89U](http://bit.ly/2dhC89U)

To identify the HTML link to scrape this data follow these steps:

1. Go to the page displaying either the PCI or PCE data
2. Right click the **Download** icon and select **Copy Link Address**
3. Paste copied link into browser window
4. Right click the **Download CSV File** icon and select **Copy Link Address**
5. Use the copied link address as the URL to scrape


```{r}
#####################
# download PCI data #
#####################
# url for PCI HTML table
url_pci <- read_html("http://www.bea.gov/iTable/iTableHtml.cfm?reqid=70&step=30&isuri=1&7022=21&7023=0&7024=non-industry&7033=-1&7025=0&7026=00000,01000,02000,04000,05000,06000,08000,09000,10000,11000,12000,13000,15000,16000,17000,18000,19000,20000,21000,22000,23000,24000,25000,26000,27000,28000,29000,30000,31000,32000,33000,34000,35000,36000,37000,38000,39000,40000,41000,42000,44000,45000,46000,47000,48000,49000,50000,51000,53000,54000,55000,56000&7027=-1&7001=421&7028=53&7031=0&7040=-1&7083=levels&7029=23&7090=70")

# download PCI table and extract the data frame from the list
pci_raw <- url_pci %>%
  html_nodes("table") %>%
  .[2] %>%
  html_table(fill = TRUE) %>%
  .[[1]]

#####################
# download PCE data #
#####################
# url for PCE HTML table
url_pce <- read_html("http://www.bea.gov/iTable/iTableHtml.cfm?reqid=70&step=10&isuri=1&7003=2&7035=-1&7004=x&7005=1&7006=00000,01000,02000,04000,05000,06000,08000,09000,10000,11000,12000,13000,15000,16000,17000,18000,19000,20000,21000,22000,23000,24000,25000,26000,27000,28000,29000,30000,31000,32000,33000,34000,35000,36000,37000,38000,39000,40000,41000,42000,44000,45000,46000,47000,48000,49000,50000,51000,53000,54000,55000,56000&7036=-1&7001=62&7002=6&7090=70&7007=-1&7093=levels")

# download PCE table and extract the data frame from the list
pce_raw <- url_pce %>%
  html_nodes("table") %>%
  .[2] %>%
  html_table(fill = TRUE) %>%
  .[[1]]

```


### Creating Tidy Data

Once the basic data has been acquired we need to pre-process it to get the data into a [tidy format](http://vita.had.co.nz/papers/tidy-data.html). This includes removing punctuations, changing the income and expenditure data from character to a numeric data type, reducing the data sets to the same time period (1997-2014), making sure the common variables share the same names, and changing the data from a wide format to a long format.  Once this has been done for both the PCI and PCE data we can merge the clean data frames into one common data frame (*data_clean*) and create a new *Savings* variable.  I also remove the District of Columbia location since this is more comparable to metropolitan-level geographic areas than state-level geographic areas. We now have the data cleaned and ready to analyze.

```{r, message=FALSE, warning=FALSE}

# create tidy PCI data
pci_clean <- pci_raw %>% 
  apply(2, function(x) gsub("[[:punct:]]", "", x)) %>%
  as_tibble(.) %>%
  group_by(GeoFips, GeoName) %>%
  mutate_each(funs(as.numeric)) %>%
  ungroup() %>%
  select(Fips = GeoFips, Location = GeoName, `1997`:`2014`) %>%
  gather(Year, Income, -c(Fips, Location))


# create tidy PCE data 
pce_clean <- pce_raw %>% 
  apply(2, function(x) gsub("[[:punct:]]", "", x)) %>%
  as_tibble(.) %>%
  group_by(Fips, Area) %>%
  mutate_each(funs(as.numeric)) %>%
  ungroup() %>%
  rename(Location = Area) %>%
  gather(Year, Expenditures, -c(Fips, Location))

# create tidy merged data frame
data_clean <- pci_clean %>%
  left_join(pce_clean) %>%
  mutate(Savings = Income - Expenditures,
         Year = as.numeric(Year)) %>%
  filter(Location != "District of Columbia")

datatable(data_clean)
```


## Exploratory Data Analysis

The primary purpose of this analysis is to assess how national and state-level PCI, PCE, and savings rates have changed over time and by geographic location. Thus, we will proceed by first assessing the national-level trends and then move on to assessing state-level trends.

### National-Level Patterns

At the national-level PCI grew by 79.6% from $22,536 in 1997 to $40,471 in 2014. Expenditures (PCE), on the other hand, grew 82.5% from $20,384 in 1997 to $37,186. So although we are assessing non-inflation adjusted dollars, since 1997 the rate of growth in PCE has outpaced PCI. Figure 1 illustrates the growing trend (not surprising since inflation has not been removed) and also captures the decrease in both PCI and PCE from 2008 to 2009 due to the [Great Recession](https://en.wikipedia.org/wiki/Great_Recession).

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.height=3.5, fig.width=8}
data_clean %>%
  filter(Location == "United States") %>%
  ggplot(aes(x = Year)) +
  geom_line(aes(y = Income, group = 1), color = "darkseagreen4") +
  geom_line(aes(y = Expenditures, group = 1), color = "firebrick3") +
  geom_ribbon(aes(ymin = Expenditures, ymax = Income, group = 1), fill = "darkseagreen1", alpha = .5) +
  annotate("text", x = 2014.25, y = 40471, label = "2014 PCI: $40.5K", color = "darkseagreen4", hjust = 0) +
  annotate("text", x = 2014.25, y = 37196, label = "2014 PCE: $37.2K", color = "firebrick3", hjust = 0) +
  annotate("text", x = 2014.25, y = 38833.5, label = "2014 Savings: $3.3K", color = "darkseagreen2", hjust = 0) +
  scale_x_continuous(NULL, limits = c(1997, 2016), breaks = seq(1998, 2014, by = 4)) +
  scale_y_continuous("Current Year Dollars", labels = scales::dollar) +
  theme_minimal()
```

However, a closer look at just the savings rate ($Savings Rate = \frac{Savings}{Income}$) depicted in Figure 2 illustrates that no consistent trend has been established. In other words, the aggregate per capita savings rate has not consistently increased or decreased year-over-year. Although a decreasing trend is present from 10% in 1998 to 6.5% in 2005, the national-level savings rate has since been fickle, peaking at 10.9% in 2012 before dipping back down to about 8-9% in recent years.  Bottom-line is that the national-level per capita savings rate has ranged between 6.5% and 10.9% with an average of 8.6%.

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.height=3.5, fig.width=8}
data_clean %>%
  filter(Location == "United States") %>%
  mutate(Savings_Rate = Savings / Income) %>%
  ggplot(aes(Year, Savings_Rate)) +
  geom_line() +
  geom_hline(aes(yintercept = mean(Savings_Rate)), linetype = "dashed", alpha = .5) +
  scale_y_continuous(NULL, labels = scales::percent, limits = c(0, .115)) +
  scale_x_continuous(NULL, breaks = seq(1998, 2014, by = 4)) +
  labs("Figure 2: National-level Savings Rate") +
  theme_minimal()
```

However, understanding aggregate ratios and trends provides limited insight regarding lower-level activity [^cost_curve]. Consequently, next we turn to investigating state-level trends. 

### State-Level Patterns

To get a quick understanding of how U.S. states have progressed over the years we can map the savings rates over time. Figure 3 highlights a few attributes:

1. Note how the earlier years have less diverging colors suggesting that there was more "equality" in the savings rates across the states; however, the latter years appear to have more disparity in the savings rates
2. As the years have progressed it appears that a growth in savings rates has been concentrated in the central states; primarily from the Dakotas down to Texas
3. A few individual states stand out:
    - Main, Vermont & Montana for savings rates that are consistently less than 0%
    - Massachusetts for consistently being a top savings rate state


```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=9, fig.height=7}
data_clean %>%
  mutate(region = tolower(Location),
         Savings_Rate = Savings / Income) %>%
  right_join(map_data("state")) %>% 
  select(-subregion) %>% 
  filter(Year %in% seq(1998, 2014, by = 2)) %>%
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Savings_Rate)) +
  facet_wrap(~ Year, ncol = 3) +
  scale_fill_gradient2(name="Savings Rate", labels = scales::percent) +
  ggtitle("Figure 3: Savings rate changes over time",
       subtitle = "Temporal map assessment of state-level savings rates (1998-2014)") +
  expand_limits() +
  theme_void() +
  theme(strip.text.x = element_text(size = 12),
        text = element_text(family = "Georgia"),
        plot.title = element_text(size = 28, margin = margin(b = 10)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)))
```

A closer look at the state-level trends provides more insight. We can see that the average saving rate over time has remained around 7%; however, confirming our assessment of the map it appears that the variance (or disparity in savings rates) has increased in recent years.  Moreover, the trend lines illustrate that with a few exceptions, states that are leading the way as top or bottom savings rate states have, historically, always been near the top or bottom. However, this should not be too surprising as it takes decades for states to change their industrial and economic infrastructure. 

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=9, fig.height=5}
savings_rate <- data_clean %>%
  mutate(Savings_Rate = Savings / Income) %>%
  filter(Location != "United States")

top5 <- savings_rate %>%
  arrange(desc(Savings_Rate)) %>%
  filter(Year == 2014) %>%
  slice(1:5)

bottom5 <- savings_rate %>%
  arrange(Savings_Rate) %>%
  filter(Year == 2014) %>%
  slice(1:5)

avg <- savings_rate %>%
  group_by(Year) %>%
  summarise(Avg_mn = mean(Savings_Rate),
            Avg_md = median(Savings_Rate)) %>%
  mutate(Avg = "Average")

ggplot(savings_rate, aes(Year, Savings_Rate, group = Location)) +
  geom_line(alpha = .1) +
  geom_line(data = filter(savings_rate, Location %in% top5$Location),
            aes(Year, Savings_Rate, group = Location), color = "dodgerblue") +
  geom_line(data = filter(savings_rate, Location %in% bottom5$Location),
            aes(Year, Savings_Rate, group = Location), color = "red") +
  geom_line(data = avg, aes(Year, Avg_mn, group = 1), linetype = "dashed") +
  annotate("text", x = 2014.25, y = .071, label = "Average", hjust = 0, size = 3) +
  geom_text_repel(data = top5, aes(label = Location), nudge_x = .5, size = 3) +
  geom_point(data = top5, aes(Year, Savings_Rate), color = "dodgerblue") +
  geom_text_repel(data = bottom5, aes(label = Location), nudge_x = 0.5, size = 3) +
  geom_point(data = bottom5, aes(Year, Savings_Rate), color = "red") +
  scale_x_continuous(NULL, limits = c(1997, 2015.25), breaks = seq(1998, 2014, by = 2)) +
  scale_y_continuous(NULL, labels = scales::percent) +
  ggtitle("Figure 4: Savings rate changes over time",
          subtitle = "Temporal assessment of state-level savings rates (1997-2014)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        text = element_text(family = "Georgia"),
        plot.title = element_text(size = 28, margin = margin(b = 10)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)))
```

However, we can also look at those states that have had the largest change in their savings rate since 1997. As Table 2 displays, three of the four states with the largest change in their savings rate were Wyoming, Oklahoma and North Dakota; all having a savings rate increase by nearly, or more than, 10%. The remaining states with the largest changes have all experienced declining savings rates, led by Nevada. 

```{r}
largest_change <- savings_rate %>%
  filter(Year == 1997 | Year == 2014) %>%
  select(Location, Year, Savings_Rate) %>%
  spread(Year, Savings_Rate) %>%
  mutate(Change = `2014` - `1997`) %>%
  arrange(desc(abs(Change))) %>%
  mutate(`1997` = paste0(round(`1997` * 100, 1), "%"),
         `2014` = paste0(round(`2014` * 100, 1), "%"),
         Change = paste0(round(Change * 100, 1), "%")) %>%
  slice(1:10)

knitr::kable(largest_change, caption = 'Table 2: States with the largest savings rate change since 1997', 
             align = c('l', 'r', 'r', 'r'))
```

This may lead us to wonder if one component (PCI vs PCE) is driving the changes in savings rate. In other words, for those states that are growing above the average level, is their PCI level growing at a greater level than those states below the average?  Or could it be that the those states with above average savings rate are experiencing a slower increase in their expenditures than those states below average. Figure 5 helps to illustrate this and suggests that...

create a boxplot that facets by Exp_Chg (from 1997-2014) and Inc_Chg and contains two boxplots in each facet that compares the rate of change for above average states vs below average states.



[^cost_curve]: The ecological fallacy and Simpson's paradox both discuss concerns of inferring the nature of lower-level trends and probabilities based on interpretation of aggregate level statistics. You can read more at: [Clark, W. A., & Avery, K. L. (1976). The effects of data aggregation in statistical analysis. *Geographical Analysis, 8*(4), 428-438.](http://onlinelibrary.wiley.com/doi/10.1111/j.1538-4632.1976.tb00549.x/abstract); [Wagner, C. H. (1982). Simpson's paradox in real life. *The American Statistician, 36*(1), 46-48](http://www.tandfonline.com/doi/abs/10.1080/00031305.1982.10482778?journalCode=utas20); [Garrett, T. A. (2003). *Aggregated versus disaggregated data in regression analysis: implications for inference. Economics Letters, 81*(1), 61-65.](http://www.sciencedirect.com/science/article/pii/S0165176503001496); [Boehmke, B. C., Johnson, A. W., White, E. D., Weir, J. D., & Gallagher, M. A. (2015). *Bending the cost curve: Moving the focus from macro-level to micro-level cost trends with cluster analysis. Journal of Cost Analysis and Parametrics, 8*(2), 126-148.](http://www.tandfonline.com/doi/abs/10.1080/1941658X.2015.1064046).
